---
title: "Explorartory Data Excercse Project 1"
author: "Curtis ONeal"
date: "May 3, 2016"
output: word_document
---

The following report outlines Explorartory Data Excercse (EDA) of the example Diamonds Dataset, specifically looking the relationship between the weight of the diamonds in carats to their price, without using a linear model to describe that relationship.

The report will follow this general outline: R code will be provided, with output as tables or charts, and a narrative passage of observations will follow, or will be interspersed with the output as comments.

#Dataset Description
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load code library dependencies, and set up data set.
library(ggplot2)
library(gridExtra)
library(magrittr)
library(scales)

#View(diamonds) #Used to check work, but doesn't show in knitted reports
pf <- diamonds #shorter data frame alias
names(pf) #get var names
pf<-pf[,-10] #remove 'z'
pf<-pf[,-9] #remove 'y'
pf<-pf[,-8] #remove 'x'
pf<-pf[,-6] #remove 'table'
#View(pf) #Used to check work, but doesn't show in knitted reports
```

The Diamonds data set has the following Variables:
"carat"   "cut"     "color"   "clarity" "depth" and "price." 

These unused variables were removed:"table" "x" "y" "z" 

```{r Explore_Vars}
#The Strucutre command tells us the number of obsevations, and details about what the variables are, their kinds, and some sample values.
str(pf)

```
The Diamonds data set has 53,940 obs (rows). After removing 4 unused variables (colums), it has 6 remaining.

The Price variable is a numeral in Dollars. It is seen as dependent on the other numerical and ordered-categorical factors of the other variables, that describe the qualities of the diamond. The categorical values are seen as having a range of qualities that get better on a scale - so they are ordered.

The following are the ordered categorical (factor) variables, with the number of their categories (levels): 
Cut (5), Clarity(7), and Color(8). 

Depth and Carat, are numerical variables.

Each observation is a diamond with each of these graded factors listed in columns. This exploration looks for a relationship in the set between weight as Carat, and Price, and a general distribution of the factors in the dataset.



#Examining Price:
```{r Explore_Price}
#This code produces a summary of diamond Prices
summary(pf$price)

#This code produces a historgram of diamond Prices
qplot(x =price, data=pf, main = 'Histogram of Diamond Prices')+ scale_x_continuous(breaks = seq(0, 18820, 2000))

#This code produces a historgram of diamond Prices
qplot(x = price, data= pf, binwidth = .1, geom = 'freqpoly', main = 'Frequency of Prices')+
  scale_x_continuous(lim = c(0,18500), breaks = seq(0, 18500, 2000)) +
  scale_y_continuous(limits = c(0,NA))

#This code produces a historgram of diamond Prices
qplot(x = price, data= pf, binwidth = .1, geom = 'freqpoly', main = 'Frequency of Prices, focusing on gap at $1,500')+
  scale_x_continuous(lim = c(0,2000), breaks = seq(0, 2000, 200)) +
  scale_y_continuous(limits = c(0,NA))
```

The price histogram is right skewed with the largest number of samples in the "$1,000 and under," category. The average price is $3,933 dollars, and the Median (middle) price is $2,401. The prices in the set range from $326 to $18,820.

In Frequency distribution shows an odd gap of missing values between $1,450 and $1,550 that should cause us to examine the data further.

#Mising Price Frequencies
```{r Plot_Price_Freq5}

#Collect, and sort ascending the unique values of price between $1,400 and $1,600
p<-sort(unique(subset(pf$price,  (pf$price < 1600 & pf$price > 1400), drop = TRUE)))
p

#Put these into a vector
price_missing <-  subset(pf$price,  (pf$price < 1600 & pf$price > 1400), drop = TRUE)
#Table was not useful 

#Create a histogram of these valuesof price between $1,400 and $1,600                 
qplot(price_missing, main="Histogram of Prices Missing between 1,454 $1,546")


```

A gap exists for values of price between $1,454 and $1,546 inclusive.   


#Examining Carat:
```{r Plot_Carat}
#"How many values does Carat take?"
length(unique(pf$carat))

#A descriptive statistical sumary of the values 
summary(pf$carat)
#table(pf$carat) #Produces a table of counts by the 273 values

#What values does Carat take?"
#sort(unique(pf$carat))
#This commented out code produces a sorted list of the 273 unique values that are too many - to view or to make it a factor.

#This code produces a historgram of Carat Values
qplot(x =carat, data=pf, main = 'Histogram of Diamond Weights in Carats') + scale_x_continuous(breaks = seq(0, 5.5, .5))
#A histogram of the carat values. Note, although there are a few in the 5.0 #range there are too few to see at this scale.

#This code produces Frequency plots of Carat Values
qplot(x = carat, data= pf, binwidth = .1, geom = 'freqpoly', main = 'Frequency of Carat Sizes, Limited to 5.5 Carats')+
  scale_x_continuous(lim = c(0,5.5), breaks = seq(0, 5.5, .1))

qplot(x = carat, data= pf, binwidth = .1, geom = 'freqpoly', main = 'Frequency of Carat Sizes, Limited to 2 Carats')+
  scale_x_continuous(lim = c(0,2), breaks = seq(0, 2, .1))

qplot(x = carat, data= pf, binwidth = .1, geom = 'freqpoly', main = 'Frequency of Carat Sizes, Limited to 2 Carats')+
  scale_x_continuous(lim = c(0,1.5), breaks = seq(0, 5.5, .1)) +
  scale_y_continuous(limits = c(0,NA), breaks = seq(0, 10500, 200))
# May need to use zoom for .8

```


In the histogram we see that Carat weights are continuos number ranges and are not confined by convention to a set series of increments such as to the .2, or .5. There are too many in the data set to create a number of factors - for other tablulation summaries.

The summary statistics are given above in the above code. Carats range from .2 to 5.01, with a median of .7, and a mean of .7979.

The carat distribution is non-normal and skewed slightly right. The long tail in the r carat range is too small to see at this scale.

The most numerous range of the sample is in the .2 to .5 carat range. But 50% of the sample is between the range of 0.4 to 1.04 Carats.

#Examining Price and Carat Together:
```{r Plot_Price_by_Carat}

#This code produces a scatter plot of Prices by Carat
qplot(x =carat, y=price, data=pf, main = 'Diamond Prices by Carat') +
  scale_x_continuous(breaks = seq(0, 5.5, .5)) +
  scale_y_continuous(breaks = seq(0, 18820, 2000))

```

The above is a scatter plot of Prices by Carat, and a general linear relationship can be seen as carat increases typically price does also. 

Many samples are outside of a linear fit line, producing a wide path, suggesting these other factors not inlcuded in this plot affect the price ranking.

There is a noticable pattern at the 1.0, 1.5, and 2.0 lines suggesting that diamonds are cut for these as minimum sizes.

A linear regression would be run by the following code, but is not assigned in this report:
CaratFirst <- lm(pf$carat~pf$price)
PriceFirst <- lm(pf$price~pf$carat)

#Examining Clarity
```{r Summary_Clarity}

#This code gives the summary statistics for Clarity
summary(pf$clarity)

#This code gives a Boxplot of Clarity values
qplot(x = clarity, y= price, 
      data= pf, 
      geom = "boxplot", main = 'Diamond Boxplot by Clarity') +
      scale_y_continuous(limits = c(0, 19000))
```
 Clarity Category Values Ordered Worst to best --> 
I1   SI2   SI1   VS2    VS1   VVS2  VVS1    IF 
741  9194  13065 12258  8171  5066  3655    1790 

Each category has the above counts in the dataset.

#Examining Depth
```{r Plot_Depth}
#This code gives the summary statistics for Depth
summary(pf$depth)

#This code gives the Histogram of Prices for Depth
qplot(x =depth, data=pf, main = 'Histogram Diamond Prices by Depth') + scale_x_continuous(breaks = seq(40, 79, 2))

#This code produces a scatterplot for Prices by Depth
qplot(x =depth, y=price, data=pf, main = 'Scatterplot of Diamond Prices by Depth') +
  scale_x_continuous(breaks = seq(40, 79, 2)) +
  scale_y_continuous(breaks = seq(0, 18820, 2000))

# Depth seems to have no real affect on price.
```

## Depth Values are Numeric Values

Depth is normally distributed centering on value 62, with 50% of the values between value 61 and 62.5.

From the plot we see that Depth seems to have almost no affect on price, since a diamond can have any depth value and be at the height or bottom of the price range.

It would be possible to creat a plot of all 273 incremental carat values in the sample, but this would not tell use very much that is useful.

#Examining Color
```{r Summary_Color}
#This code gives the summary statistics for Color
summary(pf$color)
```

Color Category Values Ordered Worst to Best --> 
D     E     F     G      H     I     J 
6775  9797  9542  11292  8304  5422  2808  

Each category has the above counts in the dataset.

#Examining Color and Price
```{r Plot_Price_Freq_Box2}

#This code gives a boxplot of Color values
qplot(x = color, y= price, 
      data= pf, 
      geom = "boxplot", main = 'Diamond Boxplot by Color') +
      scale_y_continuous(limits = c(0, 19000))

#Note the categorical value should be x, and continuous variable should be y #always true for boxplots

```

Using the Box plot of color it is easy to see that the median of the prices increase with the scale of the colors.

```{r Plot_Price_Freq_Box}
#This code gives a frequency plot of Prices, colored by Color values
qplot(x = price, data= pf, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices by Color')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
```

Since this frequency chart is so overplotted it may be useful to see separate freequency charts of the individual "color" categories to see if they show any patterns in prices or quantity.

#Prices by Individual Colors
```{r Plot_Price_Freq2}

#The below code gives, A Frequency Plot for Each Color
select_d <- subset( pf, color == "D", select = c(color, price), drop = TRUE)
only_d<- qplot(x = price, data= select_d, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color D')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_d

select_e <- subset( pf, color == "E", select = c(color, price), drop = TRUE)
only_e<- qplot(x = price, data= select_e, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color E')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_e

select_f <- subset( pf, color == "F", select = c(color, price), drop = TRUE)
only_f<- qplot(x = price, data= select_f, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color F')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_f

select_g <- subset( pf, color == "G", select = c(color, price), drop = TRUE)
only_g<- qplot(x = price, data= select_g, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color G')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_g

select_h <- subset( pf, color == "H", select = c(color, price), drop = TRUE)
only_h<- qplot(x = price, data= select_h, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color H')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_h

select_i <- subset( pf, color == "I", select = c(color, price), drop = TRUE)
only_i<- qplot(x = price, data= select_i, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color I')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_i

select_j <- subset( pf, color == "J", select = c(color, price), drop = TRUE)
only_j<- qplot(x = price, data= select_j, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices, Color J')+
  scale_x_continuous(lim = c(0,18820), breaks = seq(0,18820, 2000))
only_j

#grid.arrange(only_d, only_e, only_f, only_g, only_h, only_i, only_j, ncol = 1, newpage = TRUE) 
#This code normally creates all the charts in a single column

```

The individual plots show that colors come in all prices, but that their scarcity roughly correseponds to the scale value, but also suggest that no specific color causes an automatic price increase.

```{r Plot_Price_Freq3}
#The below code gives, A Frequency Plot for All Colors, Zooming in to prices up to $1,000

qplot(x = price, data= pf, binwidth = 10, geom = 'freqpoly', color = color, main = 'Diamond Prices by Color, Limited to $1000')+
  scale_x_continuous(lim = c(0,1000), breaks = seq(0,1000, 500))
```

The full color graph of "Price Frequencies by Color, FLimited to $1,000" is the zoomed-in version of the full color graph. 

From this chart we can infer three main ideas:  

Although the Color feature ranked with J the best and D the lowest, these colornames/letters are found at all price ranges. For Example, a J color does not automatically mean that a Diamond is worth much more. 

Looking at the chart that is cropped to a Price of $1,000, it is easier to see that the Color ranks do correspond roughly to how common each color class is. We see that J is the least common of the Colors, and the others are roughly ranked as more numerous or more commonly found the closer to the D end of the scale. 

Finally, from the individiaul plots we can tell that the actual price range and distribution is very large for each color, althought certian colors are more common at certain price ranges the range of prices is broadly distributed in the color categories. 

#More Price to Color Frequencies
```{r Plot_Price_Freq4}
#This code provides the table of Counts for Diamonds of this Color
table(pf$color)
#This code provides the table of Sums for Diamond's Prices of this Color
by(pf$price, pf$color, sum)
 
```

The above table lists the quantities of diamonds by color, and then sums the prices for all diamonds of that color.


## Examining the Cut Variable
```{r Plot_Cut}
qplot(x = pf$cut, y= price, xlab = "cut",
      data= pf, 
      geom = "boxplot") +
  coord_cartesian(ylim = c(0, 18500))

```


Strangely the cuts, although graded on a scale from Fair to Ideal, do not seem to have a clearly positive relationship to increase in price.

In summary, several factors affect the price of a diamond in the dataset, with rare color, and size appearing as the two strongest predictors. But even the easiest relationship to see in size/carat is not necessarily fully predictive since there is considerable variation in the prices of the 2.5 - 5.0 carat price range. To determine a full understanding, a model or algorithm would need to be applied. 