---
title: "EDA Project 3B -Gapminder UD651 PS-5"
author: "Curtis ONeal"
date: "May 5, 2016"
output: pdf_document
AKA: Springboard Problem Set 3
---
*******  Gapminder dataset ******
The Gapminder website contains over 500 data sets with information about
the world's population. Your task is to continue the investigation you did at the end of Problem Set 4 or you can start fresh and choose a different
data set from Gapminder.

In your investigation, examine 3 or more variables and create 2-5 plots that make use of the techniques from Lesson 5.
*********************************

Using Indicator_BMI male ASM.xlsx

Data set is titled "BMI male, age standardized mean""

The mean BMI (Body Mass Index) of the male population, counted in kilogram per square meter; this mean is calculated as if each country has the same age composition as the world population.

Primary Source: School of Public Health, Imperial College London
http://www.imperial.ac.uk/
MRC-HPA Centre for Environment and Health
http://www.imperial.ac.uk/medicine/globalmetabolics/
Secondary Source: Uploaded to Gapminder 8/2 2011
Downloaded from http://www.gapminder.org/data/ 5-2-2016

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)
library("xlsx", lib.loc="/Library/Frameworks/R.framework/Versions/3.2/Resources/library")
library(reshape2)
library(Rmisc)

getwd()
#manually placed downloaded file into wd
dat <- read.xlsx("Indicator_BMI male ASM.xlsx", sheetName="Data")
# This is an Excel of Average Male BMI Values from 
```

A quick Tour of the Data
```{r Explore, include=TRUE}

print("Variable Names")
names(dat)
#Data is in Wide format and time series cannot be accessed.

print("Number of Countrys")
length(dat$Country)
# 199 countries, 

print("Rows vs Variables- data set is in wide format")
dim(dat)

print("199 Unique BMI Values")
unique(dat$X2008)

#twenty-nine years variables from x1980-x2008, with Country being a column

print("Adjust the Data Set to Tall format")
m_dat <- melt(dat, na.rm = FALSE, variable.name = "Year", value.name = "Ave_Male_BMI")
# Melt the Year Variables into Year, and Ave_Male_BMI, Tidy Data set

print("Now has these Rows, and 3 Variables")
dim(m_dat)
#5771  Rows
# 3 Variables

print("3 Variables Named-")
names(m_dat)
#"Country"      "Year"         "Ave_Male_BMI"

#View(m_dat)

print("Summary Statistics of Ave Male BMI")
summary(m_dat$Ave_Male_BMI)

print("For Summary Graphing, Values of BMI Rounded Down Could Be Taken")
m_dat$Floor_BMI<- floor(m_dat$Ave_Male_BMI)
summary(m_dat$Floor_BMI)
print("Giving us fewer values of BMI, that would fit on a graph")
length( unique(m_dat$Floor_BMI))
m_dat$Year <- substr(m_dat$Year, 2, 5)

#Trim the X character off the Year values x1980-x2008
print("Clean up the data, check the structure and values-")
m_dat$Year <- as.numeric(m_dat$Year)
str(m_dat)

ggplot( aes( y = Ave_Male_BMI, x = Year), data = m_dat) +
  geom_point( aes(alpha = .05) )+
  ggtitle("Vertical Strip Plot of BMI's per Year (.05 Alpha)")

ggplot( aes( x = Ave_Male_BMI, y = Year), data = m_dat) +
  geom_point( aes(alpha = .05) )+
  ggtitle("Horizontal Strip Plot of BMI's per Year (.05 Alpha)")

print("These two graphs start to show us the general change, \nbut not the effect of time for specific countries.")

ggplot( aes( y = Floor_BMI, x = Year), data = m_dat) +
  geom_point( aes(alpha = .05) )+
  ggtitle("H Strip Plot of Rounded BMI's per Year (.05 Alpha)")

print("Rounding removes the clutter but doesn't give a sense of \ntrends, rather than a general visual idea of the values.")

ggplot( aes( y = Floor_BMI, x = Year), data = m_dat, size = .2) +
  geom_point( aes(alpha = .01) )+
  ggtitle("H Strip Plot of Rounded BMI's per Year (.01 Alpha & Jitter)")+
  geom_jitter(width = 0.5, height = 0.05)

print("Transparency and jittering  doesn't provide as much of a \nsense of the rarity of particular values as exepcted.")

print("For this much data we might try the heatmap or tile used in genomic data-")
ggplot(aes(y = Country, x = Year, fill = Ave_Male_BMI),
  data = m_dat) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))+
  ggtitle("Heatmap of All Countries, to Year, Colored by BMI")

print("For this we'd need a better way to identify the countries in \na systemitized way, but we do see a general trend in increase except in \n
a rare few countrys at the lowest and highest ends.")

print("To make sense of some specific trends to a limited subset  with a random sample of \n5 countries.")
set.seed(0001)
set_Country <- sample(m_dat$Country, 5, replace = FALSE)
m_dat2 <- m_dat[m_dat$Country %in% set_Country,]

#levels(m_dat2$Country)
m_dat2<- arrange(m_dat2, Country, Ave_Male_BMI)

#View(m_dat2)
# Even with the heat map, 199 Countries are nearly impossible to plot subsetting five random
# Note output will not will change due to set.seed, this could be commented out to give a random sample of countries

ggplot(aes(y = Ave_Male_BMI, x = Year, color = Country), data = m_dat2) +
  geom_line() +
  ggtitle("5 Countries, Ave Male BMI by Year")

ggplot( aes( y = Floor_BMI, x = Year, color = Country), data = m_dat2, size = .2) +
  geom_point( aes(alpha = .01) )+
  ggtitle("H Strip Plot of Rounded BMI's per Year (.01 Alpha & Jitter)")+
  geom_jitter(width = 0.5, height = 0.8)

print("This plot is specificaly interesting due to the large jumps from \n year to year in some countries, specificaly Sudan and Gabon and Serbia.")

ggplot(aes(y = Country, x = Year, fill = Ave_Male_BMI),
  data = m_dat2) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))+
  ggtitle("Heatmap of 5 Countries, by Year, \nColored by Ave Male BMI")

print("The joump in Serbia appears more gentil in the heatmap and \nthe line graph versions.")

print("Summary  Statistics Using All Countries Values-")
summary(m_dat$Ave_Male_BMI)
# This output will change anytime you run the random # sample of countries.

# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 20.93   23.64   25.62   24.96   26.67   27.91 
print("Summary  Statistics Using Only te 5 Subset Countries-")
summary(m_dat2$Ave_Male_BMI)

#drop the 194 unused levels or they appear in tables and charts
m_dat2$Country <- droplevels(m_dat2$Country)

print("A Table by Subset Country of Summary Statistics-")
table(m_dat2$Country) 
 by(m_dat2$Ave_Male_BMI, m_dat2$Country, summary)
 
```
 
Plotting the Data Individually - redoing some of the previous plots 
with the subset countries.
```{r plot, include=TRUE} 

a <-qplot(x = Ave_Male_BMI, data= m_dat, binwidth = .5, geom = 'histogram',
           main = 'Histogram of all BMI Values in OriginalDataset') +
   scale_x_continuous( name = "Averave Male BMI -Kg/M^2", limits = c(18,35), 
                       breaks = waiver(), minor_breaks = waiver(),  
                       expand = waiver(), na.value = NA_real_,
                       trans = "identity", labels = scales::comma) +
   scale_y_continuous( name = "Count")
a


a2 <-qplot(x = Ave_Male_BMI, data= m_dat2, binwidth = .5, geom = 'histogram',
           main = 'Histogram of Subset BMI Values in OriginalDataset') +
   scale_x_continuous( name = "Averave Male BMI -Kg/M^2", limits = c(18,35), 
                       breaks = waiver(), minor_breaks = waiver(),  
                       expand = waiver(), na.value = NA_real_,
                       trans = "identity", labels = scales::comma) +
   scale_y_continuous( name = "Count")
a2
 #Plain Histogram plot of All BMI values  
 
#Removed B version

c <- boxplot(Ave_Male_BMI~Country,data=m_dat2, 
        main="Ave Male BMI, by Selected Counties,1980-2008", 
        xlab="Country", ylab="Ave BMI")

# Boxplot of BMI by Selected Counties 
# Simple Boxplot of the summary data using boxplot
# Label for Longer Countries Gets Cut off "West Bank and Gaza"
# If you save this command as a variable pringint the variable gives the 
# statistics and not the plot

d<- ggplot(m_dat2, aes(x= Country, y = Ave_Male_BMI, color= Year))+
  geom_point()+
  ggtitle("Vertical Strip Plot of the Average Mail BMI Data, by Year")
d
# Vertical "Strip" Plot of the summary data, year by color
# Auto scaling allows for longer names, but overlaps
# Why Doesn't xlab work here?

e <-ggplot(m_dat2, aes(x = Ave_Male_BMI, y= Country, color =Year), 
       main="Ave Male BMI, by Selected Counties,1980-2008", 
       xlab="Country", ylab="Ave BMI")+
              geom_point()+
  ggtitle("Horizontal Strip Plot of the Average Mail BMI Data, by Year")
e

# Hoizontal "Strip" Plot of the summary data
# Labels aren't working...
# This one is better for long Country labels
```


